---
title: "PS5-rmarkdown"
author: "Tianshi Xu"
date: "2023-02-15"
output: html_document
---

#1 Load and check data (5pt)

You first task is to do a very simple data check:

1. (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title. See https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating.

#Obviously, this rmarkdown was successfully created.

2. (2pt) Load data. How many rows/columns do we have?
```{r}
library(tidyverse)
gapminder <- read_delim("gapminder.csv.bz2")
dim(gapminder)
```


3. (2pt) Print a small sample of data. Does it look OK?
```{r}
gapminder %>% 
  sample_n(10)
```

#2 Descriptive statistics (15pt)

1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.

There are `r length(unique(gapminder$name))` countries in total. `r length(unique(gapminder$iso3))` countries have 3-letter country code. `r length(unique(gapminder$iso2))` countries have 2-letter country code.

2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are even more iso3 -codes. What is going on? Can you find it out?
  (a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?

```{r}
iso2_code <- gapminder %>%
  group_by(iso2) %>%
  summarise(n_names = n_distinct(name))

head(iso2_code)

iso2_multiple_names <- iso2_code %>%
  filter(n_names > 1) %>%
  inner_join(gapminder, by = "iso2") %>%
  select(name, iso2) %>%
  distinct()
iso2_multiple_names
```

  (b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries?
Hint: two of these entitites are CHANISL and NLD CURACAO.
```{r}
iso3_code <- gapminder %>%
  group_by(name) %>%
  summarize(num_iso3 = n_distinct(iso3)) %>%
  filter(num_iso3 > 1) %>%
  arrange(desc(num_iso3))
iso3_code
```

3. (2pt) What is the minimum and maximum year in these data?

The minimum year is `r min(gapminder$time, na.rm = TRUE)`, the maximum year is `r max(gapminder$time, na.rm = TRUE)`. 


#3 CO2 emissions (30pt)

Next, let’s analyze CO2 emissions.

1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2 and co2_PC. Which years have most missing data?

There are `r sum(is.na(gapminder$co2))` missing co2 data, `r sum(is.na(gapminder$co2_PC))` missing co2_PC data.

```{r}
gapminder %>%
  group_by(time) %>%
  summarize(missing_co2 = sum(is.na(co2)))

gapminder %>%
  group_by(time) %>%
  summarize(missing_co2_PC = sum(is.na(co2_PC)))

gapminder %>%
  group_by(time) %>%
  summarize(missing_co2 = sum(is.na(co2))) %>%
  arrange(desc(missing_co2)) %>% 
  head(1)

gapminder %>%
  group_by(time) %>%
  summarize(missing_co2_PC = sum(is.na(co2_PC))) %>%
  arrange(desc(missing_co2_PC)) %>% 
  head(1)
```

2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.

```{r}
co2_data <- gapminder %>%
  filter(!is.na(co2)) %>%
  select(name, time, co2)

countries_of_interest <- c("United States of America", "China", "India", "Germany", "Japan")

co2_data_selected_countries <- co2_data %>%
  filter(name %in% countries_of_interest)

ggplot(co2_data_selected_countries, aes(x = time, y = co2, color = name)) +
  geom_line() +
  labs(title = "Total CO2 Emissions Over Time",
       x = "Year",
       y = "CO2 Emissions (kt)")
```

# I see China had a drastic increase on CO2 emission after 2000, and Germany is the only country that has a realtively decreased CO2 emission out of the 5 chosen countries.

3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC). Make a similar plot of the same countries. What does this figure suggest?
```{r}
co2_PC_data <- gapminder %>%
  filter(!is.na(co2_PC)) %>%
  select(name, time, co2_PC)

countries_of_interest <- c("United States of America", "China", "India", "Germany", "Japan")

co2_PC_data_selected_countries <- co2_PC_data %>%
  filter(name %in% countries_of_interest)

ggplot(co2_PC_data_selected_countries, aes(x = time, y = co2_PC, color = name)) +
  geom_line() +
  labs(title = "Total CO2_PC Emissions Over Time",
       x = "Year",
       y = "CO2_PC Emissions (metric tons per capita)")
```

# This figure suggest, the United States has a dominant CO2 emission per capita, but recently China increases the most. 

4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the same as continent). Comment what do you see.
Note: just compute averages over countries and ignore the fact that countries are of different size.
Hint: Americas 2016 should be 4.80.
```{r}
avg_co2_PC_continent <- gapminder %>%
  group_by(region) %>%
  summarize(avg_co2_PC = mean(co2_PC, na.rm = TRUE))
avg_co2_PC_continent
```

# I see Europe is the contient that has the highest average CO2 emission per capita, and Africa has the lowest average CO2 emission per captia.

5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016.
Hint: it should look something along these lines:
```{r}
co2_PC_continent <- gapminder %>%
  filter(!is.na(co2_PC)) %>%
  group_by(time, region) %>%
  summarize(avg_co2_PC = mean(co2_PC))

co2_PC_continent <- co2_PC_continent %>%
  filter(time %in% c(1960, 2016))

ggplot(co2_PC_continent, aes(x = region, y = avg_co2_PC, fill = as.factor(time))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Continent", y = "Average CO2 emissions per capita (metric tons)",
       title = "Average CO2 Emissions Per Capita by Continent (1960 vs. 2016)")
```

6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume region is continent).
```{r}
data_continent <- gapminder %>%
  group_by(region, time) %>%
  filter(time == 2019)

data_continent <- data_continent %>%
  mutate(co2_per_capita = co2_PC / totalPopulation * 1000)

top_co2 <- data_continent %>%
  arrange(desc(co2_per_capita)) %>%
  group_by(region) %>%
  head(n = 3)
top_co2

bottom_co2 <- data_continent %>%
  arrange(co2_per_capita) %>%
  group_by(region) %>%
  tail(n = 3)
bottom_co2
```

#4 GDP per capita (50pt)

Let’s look at GDP per capita (GDP_PC).

1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better.
Comment what do you see there.
```{r}
gapminder_1960 <- gapminder %>%
  filter(time == 1960)

continent_size <- gapminder_1960 %>%
  group_by(region) %>%
  summarize(continent_size = sum(totalPopulation)) %>%
  pull(continent_size)

ggplot(gapminder_1960, aes(x = GDP_PC, y = lifeExpectancy, color = region, size = totalPopulation)) +
  geom_point(alpha = 1) +
  labs(title = "GDP per capita vs. Life Expectancy by Country, 1960",
       x = "GDP per capita (constant 2010 US$)",
       y = "Life Expectancy at Birth (years)",
       size = "Total Population (millions)",
       color = "Continent")
```

# I see, in 1960, higher the GDP per capita, higher the life expectancy at birth can be. Among them, the best is Europe and Oceania.

2. (4pt) Make a similar plot, but this time use 2019 data only.
```{r}
gapminder_2019 <- gapminder %>%
  filter(time == 2019)

continent_size <- gapminder_2019 %>%
  group_by(region) %>%
  summarize(continent_size = sum(totalPopulation)) %>%
  pull(continent_size)

ggplot(gapminder_2019, aes(x = GDP_PC, y = lifeExpectancy, color = region, size = totalPopulation)) +
  geom_point(alpha = 1) +
  labs(title = "GDP per capita vs. Life Expectancy by Country, 2019",
       x = "GDP per capita (constant 2010 US$)",
       y = "Life Expectancy at Birth (years)",
       size = "Total Population (millions)",
       color = "Continent")
```

3. (6pt) Compare these two plots and comment what do you see. How has world developed through the last 60 years?

# I see, within the last 60 years, Europe kept its high GDP and continue leading the life expectancy throughout the world. Americas made it to the second place after 60 yeaers development. However, Africa almost stays the samee after 60 years within little change.

4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures?
Note: here as average I mean just average over countries, ignore the fact that countries are of different size.
```{r}
le_1960 <- gapminder %>% 
  group_by(region) %>% 
  filter(time == 1960) %>% 
  summarize(avg_le_1960 = mean(lifeExpectancy, na.rm = TRUE))

le_2019 <- gapminder %>% 
  group_by(region) %>%
  filter(time == 2019) %>% 
  summarize(avg_le_2019 = mean(lifeExpectancy, na.rm = TRUE))

le_continent <- merge(le_1960, le_2019, by = "region")
le_continent
```

# Yes, the result fits with what I analyzed in the question above.

5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see.
Hint: these data (data in long form) is not the simplest to compute growth. But you may want to check out the lag() function. And do not forget to group data by continent when using lag(), otherwise your results will be messed up! See https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute.
```{r}
LE_data <- gapminder %>% 
  filter(time %in% c(1960, 2019)) %>% 
  select(name, time, region, lifeExpectancy)

LE_growth <- LE_data %>% 
  group_by(name) %>% 
  mutate(growth = lifeExpectancy - lag(lifeExpectancy)) %>% 
  filter(!is.na(growth)) %>% 
  ungroup()

continent_LE_growth <- LE_growth %>% 
  group_by(region) %>% 
  summarize(avg_growth = mean(growth)) %>% 
  arrange(avg_growth)
continent_LE_growth
```

# I see Asia has the largest life expectancy growth, and Europe has the lowest life expectancy growth. This is due to the reason that Europe had a really high life expectancy relative to the other continent in the first place, so it didn't have that much space for the further high life expectancy growth.

6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it!

```{r}
gapminder_1960 <- gapminder[gapminder$time == 1960, ]
gapminder_2019 <- gapminder[gapminder$time == 2019, ]

ggplot(gapminder_1960, aes(x = GDP_PC)) +
  geom_histogram(binwidth = 500, fill = "blue", alpha = 0.5) +
  labs(title = "Histogram of GDP Per Capita in 1960",
       x = "GDP per capita (constant 2010 US$)",
       y = "Frequency")

ggplot(gapminder_2019, aes(x = GDP_PC)) +
  geom_histogram(binwidth = 500, fill = "green", alpha = 0.5) +
  labs(title = "Histogram of GDP Per Capita in 2019",
       x = "GDP per capita (constant 2010 US$)",
       y = "Frequency")

ggplot() +
  geom_histogram(data = gapminder_1960, aes(x = GDP_PC, y = ..density..), 
                 binwidth = 500, fill = "blue", alpha = 0.5) +
  geom_histogram(data = gapminder_2019, aes(x = GDP_PC, y = ..density..), 
                 binwidth = 500, fill = "green", alpha = 0.5) +
  labs(title = "Histogram of GDP Per Capita in 1960 vs. 2019",
       x = "GDP per capita (constant 2010 US$)",
       y = "Density")
```


7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.)
Hint: check out the function rank()!
Hint2: 17 for 1960.
```{r}
gapminder %>%
  filter(time == 1960) %>%
  select(name, lifeExpectancy) %>%
  arrange(desc(lifeExpectancy)) %>%
  mutate(rank = rank(desc(lifeExpectancy))) %>%
  filter(name == "United States of America")

gapminder %>%
  filter(time == 2019) %>%
  select(name, lifeExpectancy) %>%
  arrange(desc(lifeExpectancy)) %>%
  mutate(rank = rank(desc(lifeExpectancy))) %>%
  filter(name == "United States of America")
```

8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019–what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year?
Hint: 0.0904 for 1960.
```{r}
gapminder %>%
  group_by(time) %>%
  mutate(relative_rank = rank(-lifeExpectancy) / n_distinct(name[!is.na(lifeExpectancy)])) %>%
  ungroup() %>%
  select(time, name, lifeExpectancy, relative_rank) %>%
  filter(name == "United States of America")
```

# Finally tell us how many hours did you spend on this PS.

# I spent a whole day on this PS.
